\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-09 k-planar.tex
nothing changed

We first formally define $k$-planar graphs.  For a surface $\Sigma$,  a \defin{$\Sigma$-embedded graph} $G$ is a graph with $V(G)\subset\Sigma$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.  The \defin{interior} of $f$ is the point set $\{f(t):0<t<1\}$.} in $\Sigma$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior. The \defin{faces} of $G$ are the connected components of $\Sigma\setminus \bigcup_{vw\in E(G)} vw$.
If some point $p\in\Sigma$ is a common point in the interior of two distinct edges $vw$ and $xy$, then we say that $vw$ and $xy$ \defin{cross at} $p$ and we call the pair $(\{vw,xy\},p)$ a \defin{crossing} in $G$.  Without any loss of generality we may assume that no point in $\Sigma$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by local changes that do not change the number of crossings each edge is involved in.

% A \defin{crossing} in a $\Sigma$-embedded (multi)graph $G$ is a pair $(\{vw,xy\},p)$ where $vw$ and $xy$ are distinct edges of $G$ and $p$ is a point that is in the interior of both $vw$ and $xy$.

If an embedded graph $G$  has no crossings then it is \defin{non-crossing}.  When $G$ is non-crossing and $F$ is a face of $G$, we use \defin{$V(F)$} to denote the set of vertices in $G$ on the boundary of $F$.  When $F$ has a single boundary component, the \defin{facial walk} of $F$ is the walk consisting of the edges and vertices of $G$ on the boundary of $F$ in the cyclic order they are encountered while traversing the boundary of $F$.  If  $\Sigma=\R^2$ is the Euclidean plane and each edge of $G$ is involved in at most $k$ crossings, then $G$ is a $k$-plane graph.  A graph $G$ is \defin{$k$-planar} if it is isomorphic to some $k$-plane graph.  A $0$-plane graph is a \defin{plane} graph and a $0$-planar graph is a \defin{planar} graph.

@@

For any integer $d\ge 3$, an embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in some face of $G_0$. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.  A $d$-framed graph $G$ is \defin{edge-maximal} with respect to its frame $G_0$ if, for each face $F$ of $G_0$ and each pair of distinct vertices $v,w\in V(F)$, $vw\in E(F)$ or $G$ contains an edge $vw$ whose interior is contained in $F$.

\note{DW}{The word ``interior'' may be misinterpreted when refering to the outerface (see referee's comment). Shall we just write ``contained in some face''? Do we need to change this elsewhere?}
\note{PM}{Done.  I've changed occurrences elsewhere also.  Under our definitions (in Section 2.3) the faces of $G$ are open sets so in $F$ already means in the interior of $F$.  In one instance, I changed ``in the interior or boundary of $F$'' to ``in the closure of $F$.''}

Every edge-maximal $d$-framed multigraph can be described by a $(2,d(d-3)/2)$-shortcut system applied to a plane multigraph by adding a vertex inside each face $F$ of $G_0$ adjacent to the vertices of $F$ and creating a length-$2$ shortcut between each non-adjacent pair of vertices in $V(F)$. Thus \cref{ShortcutProduct,PlanarProduct} imply that any $d$-framed multigraph $G$ is a subgraph of $H\boxtimes P\boxtimes K_{O(d^2)}$ where the treewidth of $H$ is at most $10$.  However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible and the dependence on $d$ is only linear.\footnote{A multigraph $G$ is \defin{contained} in a graph $X$ if the simple graph underlying $G$ is contained in $X$.}

@@

The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A $T$-decomposition of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags contain $v_1,\ldots,v_t$.  To show the planarity of $H$ in \cref{d_framed_product_stucture} we use the fact that a graph has a 3-simple tree-decomposition if and only if it has treewidth at most 3 and is planar~\citep{kratochvil.vaner:note}.

@@

A \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ that contains a tree rooted at each vertex of $V_0$ with the property that for each $v\in V(G)$,
\[  \min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.  \]
\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite[Lemma~17]{DJMMUW20}.  For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths whose lower endpoints belong to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$ that are not already included in tripods.

\begin{lem}
  \label{induction} The setup:
  \begin{compactenum}
    \item Let $G$ be a $d$-framed multigraph and let $G_0$ be a frame of $G$.
    \item Let $F_0$ be the outer face of $G_0$ and let $T$ be a BFS spanning forest of $G_0$ rooted at $V(F_0)$.
    \item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(v,V(F_0))=j\}$.
    \item Let $F$ be a cycle in $G_0$ whose vertices are partitioned into $k\le 3$ non-empty sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
    \begin{compactenum}
      \item $F[P_i]$ is connected; and
      \item $P_i$ has a partition $\{X_i,Y_i\}$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
    \end{compactenum}
    \item Let $N$ and $N_0$ be the subgraphs of $G$ and $G_0$ consisting only of those edges and vertices contained in the closure of the interior of the Jordan curve defined by $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
  \begin{compactenum}[(i)]
    \item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
    \item For each $x\in V(H)$, $S_x$ has a partition $\{X_x,Y_x\}$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
    \item $H$ has a $3$-simple tree-decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
  \end{compactenum}
\end{lem}

@@

\begin{compactenum}
    \item If $k=1$ then let $\tau:=v_1,\ldots,v_p$ be (the facial cycle of) an inner face of $N_0$ that contains at least one edge $v_1v_2$ of $F$ on its boundary.

    \item If $k= 2$ then let $\tau:=v_1,\ldots,v_p$ be (the facial cycle of) an inner face of $N_0$ that contains at least one edge $v_1v_2$ of $F$ on its boundary and such that $v_1\in P_1$ and $v_2\in P_2$.

    \item If $k=3$ then assign each vertex $v$ of $N_0$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $V(F)$ encountered on the path in $T$ from $v$ to the root of the tree that contains $v$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now, for each $d'\in\{4,\ldots,d\}$ and each $d'$-sided inner face $F$ of $N_0$, add $d'-3$ edges inside of $F$ to split it into $3$-sided faces.  This yields a plane supergraph $N_0'$ of $N_0$ in which each inner face has exactly three edges on its boundary. By Sperner's Lemma (see \citep{Proofs4}) there exists an inner face $\tau':=v_1,v_2,v_3$ of $N_0'$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.\footnote{Although Sperner's Lemma is usually stated for simple graphs, the proof considers a certain graph defined over the faces of $N_0'$.  The proof relies only on the fact that the number of odd-degree vertices in this graph is even (known as the \defin{Handshake Lemma}), which is true even when this graph is non-simple.}  Let $\tau:=v_1,\ldots,v_p$ be (the facial cycle of) the face in $N_0$ that contains $\tau'$.
  \end{compactenum}

@@

  Let $M$ denote the subgraph of $N_0$ containing the edges and vertices of $\overline{S}$ and the edges and vertices of $F$. By our choice of $\tau$, the graph $M$ is $2$-connected. Let $F_1,\ldots,F_m,\tau$ be the inner faces of $M$. For each $i\in\{1,\ldots,m\}$, the choice of $v_1v_2v_3$ ensures that $V(F_i)$ can be partitioned into at most three sets $P_{i,1}$, $P_{i,2}$, and $P_{i,3}$ where $P_{i,1}\subseteq S$, $P_{i,2}\subseteq P_b$, and $P_{i,3}\subseteq P_c$ for some $b,c\in\{1,\ldots,k\}$.

@@

  Since $M$ is $2$-connected and $\overline{S}$ is connected, $F_i[P_{i,j}]$ is connected, for each $j\in\{1,2,3\}$.  Let $N_i$ and $N_{0,i}$ be the subgraphs of $N$ and $N_0$ contained in the closure of the interior of $F_i$. Every inner vertex of $N_{0,i}$ is an inner vertex of $N_{0}$.  Every inner face of $N_{0,i}$ is an inner face of $N_0$ and $\tau$ is not an inner face of $N_{0,i}$.  Therefore $N_{0,i}$ has no more inner vertices than $N_0$ and has fewer inner faces than $N_{0}$.  Therefore, we can apply induction using the cycle $F_i$ and the partition $P_{i,1},P_{i,2},P_{i,3}$ of $V(F_i)$.  The result of the induction is a $H_i$-partition $\mathcal{P}_i$ of $N_i$ satisfying (i)--(iii) above. We now define our partition
  \[
    \mathcal{P}:=\{Y, P_1,\ldots,P_k\} \cup \bigcup_{i=1}^m \{P\in\mathcal{P}_i: P\cap V(M)=\emptyset\} \enspace .
  \]
  It is straightforward to verify that $\mathcal{P}$ is indeed a partition of $V(N)$ and, by definition this is an $H$-partition of $N$ for the graph $H:=N/\mathcal{P}$.  It remains to verify that $\mathcal{P}$ satisfies (i)--(iii).

@@

  It is now straightforward to verify that $\mathcal{T}$ is a tree-decomposition of $H_0:=N_0/\mathcal{P}$ in which each bag has size at most $4$.    To see that $\mathcal{T}$ is also a tree-decomposition of $H:=N/\mathcal{P}$, consider any edge $vw\in E(N)\setminus E(N_0)$ with $v\in P_\alpha$ and $w\in P_\beta$.  We must verify that some bag of $\mathcal{T}$ contains $\alpha$ and $\beta$.  If both $v$ and $w$ are in $V(M)$, then $\alpha=x_a$ and $\beta=x_b$ for some $a,b\in\{0,\ldots,3\}$, in which case $B_{y_0}=\{x_0,\ldots,x_k\}$ contains $\alpha$ and $\beta$.  If neither $v$ nor $w$ are in $V(M)$ then, since no edge of $N\subseteq G$ crosses an edge of $M\subseteq G_0$, both $v$ and $w$ are contained in the same face $F_i$ of $M$. So, by the inductive hypothesis there exists a bag of $\mathcal{T}_i$ that contains $\alpha$ and $\beta$ and this becomes a bag of $\mathcal{T}$ that contains $\alpha$ and $\beta$.  Finally, if $v\in V(M)$ and $w\not\in V(M)$, then $v\in P_{i,j}$ for some $j\in\{1,2,3\}$ and $w$ is in some face $F_i$ of $M$. By induction, some bag of $\mathcal{T}_i$ contains $\beta$ and $\alpha'$, where $\alpha'$ is the vertex of $H_i$ obtained by contracting $P_{i,j}$.  However, before $T_i$ is attached to $T_0$ each occurence of $\alpha'$ in $\mathcal{T}_i$ is replaced by $\alpha$, so some bag of $\mathcal{T}$ contains both $\alpha$ and $\beta$. Therefore $\mathcal{T}$ is a tree-decomposition of $H$.

@@

Define $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{\lfloor d/2\rfloor i}\cup \cdots \cup L_{\lfloor d/2\rfloor(i+1)-1}$ for each integer $i\ge 0$. This is a layering of $G$ since $\dist_{G_0}(v,w)\le \floor{ \frac{d}{2} }$ for each edge $vw\in E(G)$.  By Condition~(ii), $|L_j\cap Y_x|\le 3$ for each integer $j\ge 0$, so $|L_i'\cap Y_x|\le 3\floor{\tfrac{d}{2}}$. Therefore, $|L_i'\cap S_x|= |L_i'\cap Y_x|+ |L_i'\cap X_x|\le 3\floor{ \tfrac{d}{2}} + d -3$ for each integer $i\ge 0$ and each $x\in V(H)$. The result now follows from \cref{PartitionProduct}.
\end{proof}

@@

\begin{proof}
Let $G$ be a (connected) $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}. We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If distinct vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no newly-introduced edge $vw$ is on the boundary of a face with an edge parallel to $vw$.  Since $G$ is connected, each edge added during this process splits a face $F$ with a facial walk of length at least $4$ into two faces each having a shorter facial walk than that of $F$.
Therefore this process eventually reaches a stage where each face is either bounded by a $3$-cycle or a (pre-existing) $2$-cycle and the process ends.

@@

To understand the structure of $G$, it is helpful to consider the plane graph $G'$  obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G'$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge-maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$. If $W$ contains a dummy vertex $w_i$, then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) would be involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is flanked by two non-dummy vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4>1$ in $G'$ and no two edges incident to a common vertex cross each other, $w_{i-1}\neq w_{i+1}$.  Since $G$ is edge-maximal, the edge $w_{i-1}w_{i+1}$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex $w_i$ in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G'$, each of which is contained in a different edge of $G$.

@@

If $G_0$ has a face $F$ of length 2, then add a new vertex inside $F$ adjacent to both vertices on the boundary of $F$, which creates two new triangular faces $F_1$ and $F_2$. If $F$ is a lake, then make $F_1$ and $F_2$ lakes. If $F$ is a nation, then make $F_1$ a nation and make $F_2$ a lake. The resulting map graph is still $G$. We may therefore assume that $G_0$ has no face of length $2$.  We may also assume that $G_0$ is edge-maximal in the following sense: Adding any edge to $G_0$ will introduce a crossing or make it impossible to label the faces of $G_0$ as nations or lakes to obtain $G$ as the resulting map graph. This edge-maximal graph is well-defined since the assumption of having no face of length $2$ implies that $|E(G_0|\leq 3|V(G_0)|-6$.

We now argue that $G_0$ is biconnected.  Suppose that some face $F$ of $G_0$ has a disconnected boundary (implying that $G_0$ is disconnected). Let $v$ and $w$ be vertices in distinct components of the boundary of $F$. Add an edge $vw$ to $G_0$ inside of $F$. The corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no face of $G_0$ has a disconnected boundary and therefore $G_0$ is connected.

>>>>

We first formally define $k$-planar graphs.  For a surface $\Sigma$,  a \defin{$\Sigma$-embedded graph} $G$ is a graph with $V(G)\subset\Sigma$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.  The \defin{interior} of $f$ is the point set $\{f(t):0<t<1\}$.} in $\Sigma$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior. The \defin{faces} of $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$. A \defin{crossing} in a $\Sigma$-embedded (multi)graph $G$ is an unordered pair of distinct edges $\{vw,xy\}\subseteq E(G)$ whose interiors have a point in common.

If an embedded graph $G$  has no crossings then it is \defin{non-crossing} graph.  When $G$ is non-crossing and $F$ is a face of $G$, then let \defin{$V(F)$} be the set of vertices in $G$ on the boundary of $F$.  When $F$ is homeomorphic to a disk, the \defin{facial walk} of $F$ is the cycle consisting of the edges and vertices of $G$ on the boundary of $F$.  If  $\Sigma=\R^2$ is the Euclidean plane and each edge of $G$ is involved in at most $k$ crossings, then $G$ is a $k$-plane graph.  A graph $G$ is \defin{$k$-planar} if is isomorphic to some $k$-plane graph.  A $0$-plane graph is a \defin{plane} graph and a $0$-planar graph is a \defin{planar} graph.

Without any loss of generality we may assume that the interiors of any two edges of a $\Sigma$-embedded graph $G$ have at most one point in common and that no point in $\R^2$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by local changes that do not change the number of crossings each edge is involved in.  If $p$ is a common point in the interior of two crossing edges $vw$ and $xy$, then we say that $vw$ and $xy$ \defin{cross at} $p$.

@@

For any integer $d\ge 3$, an embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$ \note{DW}{The word ``interior'' may be misinterpreted when refering to the outerface (see referee's comment). Shall we just write ``contained in some face''? Do we need to change this elsewhere?}. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.

Every $d$-framed multigraph can be described by a $(2,d(d-3)/2)$-shortcut system applied to a plane multigraph by adding a vertex inside each face $F$ of $G_0$ adjacent to the vertices of $F$ and creating a shortcut between each non-adjacent pair of vertices in $V(F)$. Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible.\footnote{A multigraph $G$ is \defin{contained} in a graph $X$ if the simple graph underlying $G$ is contained in $X$.}

@@

\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite[Lemma~17]{DJMMUW20}. The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A $T$-decomposition of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags contain $v_1,\ldots,v_t$.  To show the planarity of $H$ in \cref{d_framed_product_stucture} we use the fact that a graph has a 3-simple tree-decomposition if and only if it has treewidth at most 3 and is planar~\citep{kratochvil.vaner:note}. Finally, a \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ that contains a tree rooted at each vertex of $V_0$ with the property that for each $v\in V(G)$,
\[  \min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.  \]

@@

For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths belonging to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$.

\begin{lem}
  \label{induction} The setup:
  \begin{compactenum}
    \item Let $G$ be a $d$-framed multigraph and let $G_0$ be a frame of $G$.
    \item Let $F_0$ be the outer face of $G_0$ and let $T$ be a BFS spanning forest of $G_0$ rooted at $V(F_0)$.
    \item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(v,V(F_0))=j\}$.
    \item Let $F$ be a cycle in $G_0$ whose vertices are partitioned into $k\le 3$ non-empty sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
    \begin{compactenum}
      \item $F[P_i]$ is connected; and
      \item $P_i$ has a partition $\{X_i,Y_i\}$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
    \end{compactenum}
    \item Let $N$ and $N_0$ be the subgraphs of $G$ and $G_0$ consisting only of those edges and vertices contained in the interior or boundary of $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
  \begin{compactenum}[(i)]
    \item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
    \item For each $x\in V(H)$, $S_x$ has a partition $\{X_x,Y_x\}$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
    \item $H$ has a $3$-simple tree-decomposition $\mathcal{T}$, and if $k=3$ then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
  \end{compactenum}
\end{lem}

@@

\begin{compactenum}
    \item If $k=1$ then let $\tau:=v_1,\ldots,v_p$ be an inner face of $N_0$ that contains at least one edge $v_1v_2$ of $F$ on its boundary.

    \item If $k= 2$ then let $\tau:=v_1,\ldots,v_p$ be an inner face of $N_0$ that contains at least one edge $v_1v_2$ of $F$ on its boundary and such that $v_1\in P_1$ and $v_2\in P_2$.

    \item If $k=3$ then assign each vertex $v$ of $N_0$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $V(F)$ encountered on the path in $T$ from $v$ to the root of the tree that contains $v$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now, for each $d'\in\{4,\ldots,d\}$ and each $d'$-sided inner face $F$ of $N_0$, add $d'-3$ edges inside of $F$ to split it into $3$-sided faces.  This yields a plane supergraph $N_0'$ of $N_0$ in which each inner face has exactly three edges on its boundary. By Sperner's Lemma (see \citep{Proofs4}) there exists an inner face $\tau':=v_1,v_2,v_3$ of $N_0'$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.\footnote{Although Sperner's Lemma is usually stated for simple graphs, the proof considers a certain graph defined over the faces of $N_0'$.  The proof relies only on the fact that the number of odd-degree vertices in this graph is even (known as the \defin{Handshake Lemma}), which is true even when this graph is non-simple.}  Let $\tau:=v_1,\ldots,v_p$ be  the facial cycle of $N_0$ that contains $\tau'$.
  \end{compactenum}

@@

  Let $M$ denote the subgraph of $N_0$ containing the edges and vertices of $\overline{S}$ and the edges and vertices of $F$. By our choice of $\tau$, the graph $M$ is $2$-connected. Let $F_1,\ldots,F_m,\tau$ be the inner faces of $M$. By our choice of $v_1v_2v_3$, the vertices of each $F_i$ can be partitioned into at most three sets $P_{i,1}$, $P_{i,2}$, and $P_{i,3}$ where $P_{i,1}\subset Y$, $P_{i,2}\subseteq P_b$, and $P_{i,3}\subseteq P_c$ for some $b,c\in\{1,\ldots,k\}$.

@@

  Since $M$ is $2$-connected and $\overline{S}$ is connected, $F_i[P_{i,j}]$ is connected, for each $j\in\{1,2,3\}$.  Let $N_i$ and $N_{0,i}$ be the subgraphs of $N$ and $N_0$ contained in the interior or boundary of $F_i$.  Every inner vertex of $N_{0,i}$ is an inner vertex of $N_{0}$.  Every inner face of $N_{0,i}$ is an inner face of $N_0$ and $\tau$ is not an inner face of $N_{0,i}$.  Therefore $N_{0,i}$ has no more inner vertices than $N_0$ and has fewer inner faces than $N_{0}$.  Therefore, we can apply induction using the cycle $F_i$ and the partition $P_{i,1},P_{i,2},P_{i,3}$ of $V(F_i)$.  The result of the induction is a $H_i$-partition $\mathcal{P}_i$ of $N_i$ satisfying (i)--(iii) above. We now define our partition
  \[
    \mathcal{P}:=\{Y, P_1,\ldots,P_k\} \cup \bigcup_{i=1}^m \{P\in\mathcal{P}_i: P\cap V(M)=\emptyset\} \enspace .
  \]
  It is straightforward to verify that $\mathcal{P}$ is indeed a partition of $V(N)$ and, by definition this is an $H$-partition of $N$ for the graph $H:=N/\mathcal{P}$.  It remains to verify that $\mathcal{P}$ satisfies (i)--(iii).

@@

  It is now straightforward to verify that $\mathcal{T}$ is a tree-decomposition of $H_0:=N_0/\mathcal{P}$ in which each bag has size at most $4$.    To see that $\mathcal{T}$ is also a tree-decomposition of $H:=N/\mathcal{P}$, consider any edge $vw\in E(N)\setminus E(N_0)$ with $v\in P_\alpha$ and $w\in P_\beta$.  We must verify that some bag of $\mathcal{T}$ contains $\alpha$ and $\beta$.  If both $v$ and $w$ are in $V(M)$, then $\alpha=x_a$ and $\beta=x_b$ for some $a,b\in\{0,\ldots,3\}$, in which case $B_{y_0}=\{x_0,\ldots,x_k\}$ contains $\alpha$ and $\beta$.  If neither $v$ nor $w$ are in $V(M)$ then, since no edge of $N\subseteq G$ crosses an edge of $M\subseteq G_0$, both $v$ and $w$ are contained in the interior of the same face $F_i$ of $M$. So, by the inductive hypothesis there exists a bag of $\mathcal{T}_i$ that contains $\alpha$ and $\beta$ and this becomes a bag of $\mathcal{T}$ that contains $\alpha$ and $\beta$.  Finally, if $v\in V(M)$ and $w\not\in V(M)$, then $v\in P_{i,j}$ for some $j\in\{1,2,3\}$ and $w$ is in the interior of some face $F_i$ of $M$. By induction, some bag of $\mathcal{T}_i$ contains $\beta$ and $\alpha'$, where $\alpha'$ is the vertex of $H_i$ obtained by contracting $P_{i,j}$.  However, before $T_i$ is attached to $T_0$ each occurence of $\alpha'$ in $\mathcal{T}_i$ is replaced by $\alpha$, so some bag of $\mathcal{T}$ contains both $\alpha$ and $\beta$. Therefore $\mathcal{T}$ is a tree-decomposition of $H$.

@@

Define $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{\lfloor d/2\rfloor i}\cup \cdots \cup L_{\lfloor d/2\rfloor(i+1)-1}$ for each integer $i\ge 0$. This is a layering of $G$ since $\dist_{G_0}(v,w)\le \floor{ \frac{d}{2} }$ for each edge $vw\in E(G)$.  By Condition~(ii), $|L_j\cap Y_x|\le 3$ for each integer $j\ge 0$, so $|L_i'\cap Y_x|\le 3\floor{\frac{d}{2}}$. Therefore, $|L_i'\cap S_x|= |L_i'\cap Y_x|+ |L_i'\cap X_x|\le \floor{ \frac{d}{2}} + d -3$ for each integer $i\ge 0$ and each $x\in V(H)$. The result now follows from \cref{PartitionProduct}.
\end{proof}

@@

\begin{proof}
Let $G$ be a (connected) $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}. We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If distinct vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no newly-introduced edge $vw$ is on the boundary of a face with an edge parallel to $vw$.  Since $G$ is connected, each edge added during this process splits a face $F$ with a facial walk of length at least $4$ into two faces each having a shorter facial walk than that of $F$.
Therefore this process eventually reaches a stage where each face is either bounded by a $3$-cycle or (a pre-existing) $2$-cycle and the process ends.

@@

To understand the structure of $G$, it is helpful to consider the plane graph $G'$  obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G'$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge-maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

\note{DW}{I suggest we do not start a new paragraph here}
If $W$ contains a dummy vertex $w_i$, then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) would be involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is flanked by two non-dummy vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4>1$ in $G'$ and no two edges incident to a common vertex cross each other, $w_{i-1}\neq w_{i+1}$.  Since $G$ is edge-maximal, the edge $w_{i-1}w_{i+1}$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex $w_i$ in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G'$, each of which is contained in a different edge of $G$.

@@

\note{PM}{Here we're using $f$ to denote a face, earlier we were using $F$.}
If $G_0$ has a face $f$ of length 2, then add a new vertex inside $f$ adjacent to both vertices on the boundary of $f$, which creates two new triangular faces $f_1$ and $f_2$. If $f$ is a lake, then make $f_1$ and $f_2$ lakes. If $F$ is a nation, then make $f_1$ a nation and make $f_2$ a lake. The resulting map graph is still $G$. So we may assume that $G_0$ is an edge-maximal multigraph with no faces of length 2 (and with each face labelled a nation or a lake), such that $G$ is the corresponding map graph. This is well-defined since the assumption of having no face of length 2 implies that $|E(G_0|\leq 3|V(G_0)|-6$.

Suppose that some face $f$ of $G_0$ has a disconnected boundary (implying that $G_0$ is disconnected). Let $v$ and $w$ be vertices in distinct components of the boundary of $f$. Add the edge $vw$ to $G_0$ across $f$. The corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no face of $G_0$ is disconnected.

Suppose that some face $f$ of $G_0$ has a repeated vertex $v$ in the boundary walk of $f$ and let $u,v,w$ be consecutive vertices in this boundary walk.  Since $v$ appears more than once in the boundary walk of $f$, the degree of $v$ is greater than $1$.  Since the degree of $v$ is greater than $1$ and the length of $f$ is greater than $2$, $u\neq w$. Since $v$ is repeated, $uw$ is not an edge in the boundary of $f$. Add the edge $uw$ across $f$ and consider the resulting face $uvw$ to be a lake. Since $v$ appears elsewhere in the boundary of $f$, the corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no facial walk of $G_0$ has a repeated vertex. Since each facial walk is connected, each face of $G_0$ is bounded by a cycle, and $G_0$ is biconnected.
